Sampling Strategies in Decisions from Experience

### 2022-05-05 manuscript.Rmd
Refine plots (add summary analysis)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, #omit messages
                      fig.align = "l", #align figures to left margin
                      fig.pos = "t", #display figures at the top of a page
                      fig.height=7, 
                      fig.width=10)
```

```{r packages, include=FALSE}
#load required packages
pacman::p_load(papaja,
               knitr,
               readr,
               tidyr, 
               dplyr,
               ggplot2,
               viridis,
               scico,
               ggpubr,
               latex2exp)
```

@@

```{r echo=TRUE}
# define a prospect with two outcomes
outcomes <- c(5, 0)
probabilities <- c(.2, .8)

@@

```{r sampling-paradigm, fig.cap="(ref:sampling-paradigm)"}
knitr::include_graphics(path = "manuscript_files/sampling-paradigm.png")
```

@@

Below, the key predictions concerning the effects of the model parameters $\psi$ and $\theta$ on the length of the sampling and accumulation process and the resulting choice are outlined.
Moreover, previous research has shown that differences as to how information about the properties of prospects is processed can translate to characteristic shapes of cumulative prospect theory's (CPT) value and weighting function [see @pachurHowTwainCan2017; @zilkerNonlinearProbabilityWeighting2021].
Respectively, this section also contains predictions about the relations between $\psi$ and $\theta$ and the CPT parameters.
A summary of the model predictions can be found in Table\ \@ref(tab:predictions).

@@

```{r predictions}
predictions <- tibble("Measure" = 
                        c("Choices pattern",
                          "",
                          "$\\gamma$ (curvature of $w$)",
                          "$\\delta$ (elevation of $w$)",
                          "$\\alpha$ (concavity of $v$)"),
                      "Predictions for changes in $\\psi$" = 
                        c("More EV maximization for decreasing $\\psi$",
                          "More as-if underweighting of rare outcomes for increasing $\\psi$",
                          "$\\geq 1$, increases with $\\psi$",
                          "1, no effect of $\\psi$",
                          "$\\leq 1$, decreases with increasing $\\psi$"))
apa_table(predictions, caption = "(ref:predictions)", note = "The predictions should become more robust for high values of $\\theta$.", escape = FALSE)
```

@@

```{r weighting-function, fig.cap="(ref:weighting-function)"}
#compute images of weighting function (wf)
wf <- tibble(p = seq(0, 1, .01)) %>%  #cumulative probabilities
  expand_grid(gamma = seq(.1, 2, .1), #gamma values
              delta = c(.1, .5, 1, 2, 5, 10)) %>% #delta values
  mutate(wp = (delta*(p^gamma))/((delta*p^gamma)+(1-p)^gamma)) #images of wf

@@

* *Decision variable*: The decision variable of a prospect either counts the number of won comparisons---implying an absolute threshold---or how many more comparisons were won relative to the other prospect---implying a relative threshold.^[The results are similar for absolute and relative thresholds. For simplification, only the results of the absolute thresholds are reported in the main text.]

@@

```{r data, include=FALSE}
#load choice data
cols_choices <- list(.default = col_double(),
                     boundary = col_factor(),
                     gamble = col_factor(),
                     rare = col_factor(),
                     agent = col_factor(),
                     choice = col_factor())
choices <- read_csv("data/simulation_summary.csv", col_types = cols_choices)
choices <- choices %>% 
  filter(boundary == "absolute")

#load CPT estimates
cols_cpt <- list(.default = col_double(),
                 boundary = col_factor(),
                 a = col_factor(),
                 parameter = col_factor())
cpt <- read_csv("data/cpt_parameters.csv", col_types = cols_cpt)
cpt <- cpt %>% 
  filter(boundary == "absolute")
```

@@

Overall, the results of the simulation match the predictions from Table\ \@ref(tab:predictions), however, with some exceptions concerning the estimates for the parameter $\delta$ of the weighting function. 
We start by inspecting how the length of the information-processing chain changes with varying combinations of $\psi$ and $\theta$.

@@

```{r chain-length, fig.cap="(ref:chain-length)"}
choices %>%
  group_by(s, a) %>% 
  summarize(n_med = median(n_sample)) %>% 
  ggplot(aes(x = s, y = n_med, group = as.factor(a), color = as.factor(a))) + 
   labs(x = expression(psi),
       y = "Median of Overall Sampled Outcomes",
       color = expression(theta)) +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, .5)) +
  scale_y_continuous(limits = c(0, 120), breaks = seq(0, 150, 50)) +
  geom_point() +
  geom_line() + 
  geom_hline(yintercept = 14, linetype = "dashed", color = "gray") +
  annotate(geom="text", x=.95, y=16, label="Wulff et al. (2018)", color="gray") + 
  scale_color_viridis_d(option = "plasma") +
  theme_apa()
```

@@

```{r choice-rates, fig.cap="(ref:choice-rates)"}
#prepare data

@@

(ref:weighting-parameters) Parameter Estimates and Graphical Shape of the Weighting Function

@@

#plot estimates

##gamma
gamma <- cpt %>%
  filter(parameter == "gamma") %>%
  ggplot(aes(s, mean)) +
  facet_wrap(~a, nrow = 1, labeller = labeller(a = as_labeller(label_theta, default = label_parsed))) +
  scale_y_continuous(limits = c(0,2), breaks = seq(0,2.0,1.0))+
  labs(x = element_blank(), #omit axis title
       y = expression(gamma)) +
  geom_errorbar(aes(ymin=`2.5%`, ymax=`97.5%`)) + 
  geom_point() +
  theme_apa() +
  theme(axis.line.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank())

##delta
delta <- cpt %>%
  filter(parameter == "delta") %>% 
  ggplot(aes(s, mean)) +
  facet_wrap(~a, nrow = 1) + 
  scale_y_continuous(limits = c(0,10), breaks = seq(0,10,5)) +
  scale_x_continuous(limits = c(0, 1.1), breaks = seq(0,1,.5)) +
  labs(x = expression(psi),
       y = expression(delta)) +
  geom_errorbar(aes(ymin=`2.5%`, ymax=`97.5%`)) + 
  geom_point() +
  theme_apa() + 
  theme(strip.text.x = element_blank()) # omit facet labels

wf <- cpt %>%
  select(s, boundary, a, parameter, mean) %>%
  pivot_wider(names_from = parameter, values_from = mean) %>%
  select(-c(alpha, rho)) %>%
  expand_grid(p = seq(0, 1, .05)) %>%
  mutate(w = round(  (delta * p^gamma)/ ((delta * p^gamma)+(1-p)^gamma), 2))

##weighting function
weight <- wf %>% #graphs for absolute boundary
  ggplot(aes(p, w, group = s, color = s)) +
  facet_wrap(~a, nrow = 1, labeller = labeller(a = as_labeller(label_theta, default = label_parsed))) + 
  labs(x = "Experienced Probability",
       y = expression(pi),
       color = expression(psi)) +
  scale_x_continuous(breaks = seq(0, 1, .5)) +
  scale_y_continuous(breaks = seq(0, 1, .5)) +
  geom_line(size = .5) +
  scale_color_viridis(option = "viridis") +
  theme_apa() + 
  theme(strip.text.x = element_blank()) #omit facet labels

##merge figures
ggarrange(gamma, delta, weight, ncol = 1, nrow = 3, common.legend = TRUE, legend = "right")
```

@@

(ref:value-parameters) Parameter Estimates and Graphical Shape of the Value Function

```{r value-parameters, fig.cap="(ref:value-parameters)"}
#plot estimates

##alpha
alpha <- cpt %>%
  filter(parameter == "alpha") %>% 
  ggplot(aes(s, mean)) +
  facet_wrap(~a, nrow = 1, labeller = labeller(a = as_labeller(label_theta, default = label_parsed))) +
  scale_x_continuous(limits = c(0, 1.1), breaks = seq(0,1,.5)) +
  scale_y_continuous(limits = c(0,1), breaks = seq(0,1, .5)) + 
  labs(x = element_blank(),
       y = expression(alpha),
       color = expression(psi)) +
  geom_point() +
  geom_errorbar(aes(ymin=`2.5%`, ymax=`97.5%`)) + 
  theme_apa() + 
  theme(axis.line.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank())

###value function
vf <- cpt %>%
  select(s, boundary, a, parameter, mean) %>%
  pivot_wider(names_from = parameter, values_from = mean) %>%
  select(-c(gamma, delta, rho)) %>%
  expand_grid(x = seq(0, 20, .5)) %>%  
  mutate(v = round(x^alpha, 2))

value <- vf %>% 
  ggplot(aes(x, v, group = s, color = s)) +
  facet_wrap(~a, nrow = 1)+ 
  labs(x = "Objective Outcome",
       y = "Subjective Value",
       color=expression(psi)) +
  scale_x_continuous(limits = c(0,20), breaks = seq(0, 20, 10)) +
  scale_y_continuous(limits = c(0,20), breaks = seq(0, 20, 10)) +
  geom_line(size = .5) +
  scale_color_viridis(option = "viridis") +
  theme_apa() + 
  theme(strip.text.x = element_blank())

##rho
rho <- cpt %>%
  filter(parameter == "rho") %>% 
  ggplot(aes(s, mean)) +
  facet_wrap(~a, nrow = 1) + 
  labs(x = expression(psi),
       y = expression(rho),
       color = expression(psi)) +
  scale_x_continuous(limits = c(0, 1.1), breaks = seq(0, 1,.5)) +
  scale_y_continuous(limits = c(0, 5), breaks = seq(0, 5,2.5)) +
  geom_point() +
  geom_errorbar(aes(ymin=`2.5%`, ymax=`97.5%`)) + 
  theme_apa() + 
  theme(strip.text.x = element_blank())

##merge figure
ggarrange(alpha, rho, value, ncol = 1, nrow = 3, common.legend = TRUE, legend = "right") 
```

# Summary and Discussion

@@

```{r}
#load CPT estimates
cols_cpt <- list(.default = col_double(),
                               boundary = col_factor(),
                               a = col_factor(),
                               parameter = col_factor())
cpt_c <- read_csv("supplements/estimates_cpt_comprehensive.csv", col_types = cols_cpt)
cpt_c <- cpt_c %>% 
  filter(boundary == "absolute")

cpt <- cpt %>% mutate(integration = "round")
cpt_c <- cpt_c %>% mutate(integration = "summary")
cpt_both <- bind_rows(cpt, cpt_c)
```

```{r test}

#plot estimates

gamma_2 <- cpt_both %>%
  filter(parameter == "gamma") %>%
  ggplot(aes(s, mean, color = integration)) +
  facet_wrap(~a, nrow = 2, labeller = labeller(a = as_labeller(label_theta, default = label_parsed)), scales = "free_y") +
  scale_x_continuous(limits = c(0,1), breaks = seq(0,1,.5)) + 
  scale_color_scico_d(palette = "berlin") + 
  labs(x = expression(psi), 
       y = expression(gamma)) +
  geom_errorbar(aes(ymin=`2.5%`, ymax=`97.5%`)) + 
  geom_point() +
  geom_line() + 
  theme_apa()

delta_2 <- cpt_both %>%
  filter(parameter == "delta") %>%
  ggplot(aes(s, mean, color = integration)) +
  facet_wrap(~a, nrow = 2, labeller = labeller(a = as_labeller(label_theta, default = label_parsed)), scales = "free_y") +
  scale_x_continuous(limits = c(0,1), breaks = seq(0,1,.5)) + 
  scale_color_scico_d(palette = "berlin") + 
  labs(y = expression(delta)) +
  geom_errorbar(aes(ymin=`2.5%`, ymax=`97.5%`)) + 
  geom_point() +
  geom_line() + 
  theme_apa()

ggarrange(gamma_2, delta_2, ncol = 1, nrow = 2, common.legend = TRUE, legend = "right")
```

>>>>

```{r setup, include=FALSE}
# set global chunk options
knitr::opts_chunk$set(echo = FALSE, 
                      fig.align = "l", 
                      fig.pos = "t", 
                      fig.height=7, 
                      fig.width=10)
```

```{r packages, include=FALSE}
# load required packages
pacman::p_load(papaja,
               tidyverse,
               viridis, 
               scico, 
               ggpubr,  
               latex2exp 
               )
```

```{r data, include=FALSE}
# load and prepare required data

## choice data

### specify column types
cols_choices <- list(.default = col_double(),
                     boundary = col_factor(),
                     gamble = col_factor(),
                     rare = col_factor(),
                     agent = col_factor(),
                     choice = col_factor())

### load data files from round-wise and summary model
choices_roundwise <- read_csv("data/choices_roundwise.csv", col_types = cols_choices)
choices_summary <- read_csv("data/choices_summary.csv", col_types = cols_choices)

### merge data files 
choices_roundwise <- choices_roundwise %>%
  mutate(model = "roundwise") %>% 
  select(model, everything())

choices_summary <- choices_summary %>%
  mutate(model = "summary") %>% 
  select(model, everything())

choices <- bind_rows(choices_roundwise, choices_summary)

## cpt estimates
cols_cpt <- list(.default = col_double(),
                 boundary = col_factor(),
                 a = col_factor(),
                 parameter = col_factor())

cpt_roundwise <- read_csv("data/cpt-parameters_roundwise.csv", col_types = cols_cpt)
cpt_summary <- read_csv("data/cpt-parameters_summary.csv", col_types = cols_cpt)

cpt_roundwise <- cpt_roundwise %>%
  mutate(model = "roundwise") %>% 
  select(model, everything())

cpt_summary <- cpt_summary %>%
  mutate(model = "summary") %>% 
  select(model, everything())

cpt <- bind_rows(cpt_roundwise, cpt_summary)
```

@@

```{r code-example-1, echo=TRUE}
# define a prospect with two outcomes
outcomes <- c(5, 0)
probabilities <- c(.2, .8)

@@

```{r fig-sampling-paradigm, fig.cap="(ref:sampling-paradigm)"}
knitr::include_graphics(path = "manuscript_files/sampling-paradigm.png")
```

@@

Below, the key predictions concerning the effects of the model parameters $\psi$ and $\theta$ on the length of the sampling and accumulation process and the resulting choice are outlined.
Moreover, previous research has shown that differences as to how information about the properties of prospects is processed can translate to characteristic shapes of cumulative prospect theory's (CPT) value and weighting function [see @pachurHowTwainCan2017; @zilkerNonlinearProbabilityWeighting2021].
Respectively, this section also contains predictions about the relations between $\psi$ and $\theta$ and the CPT parameters.
A summary of the model predictions can be found in Table\ \@ref(tab:tab-predictions).

@@

```{r tab-predictions}
predictions <- tibble("Measure" = 
                        c("Choices pattern",
                          "",
                          "$\\gamma$ (curvature of $w$)",
                          "$\\delta$ (elevation of $w$)",
                          "$\\alpha$ (concavity of $v$)"),
                      "Predictions for changes in $\\psi$" = 
                        c("More EV maximization for decreasing $\\psi$",
                          "More as-if underweighting of rare outcomes for increasing $\\psi$",
                          "$\\geq 1$, increases with $\\psi$",
                          "1, no effect of $\\psi$",
                          "$\\leq 1$, decreases with increasing $\\psi$"))
apa_table(predictions, caption = "(ref:predictions)", note = "The predictions should become more robust for high values of $\\theta$.", escape = FALSE)
```

@@

```{r fig-weighting-function, fig.cap="(ref:weighting-function)"}
#compute images of weighting function (wf)
wf <- tibble(p = seq(0, 1, .01)) %>%  #cumulative probabilities
  expand_grid(gamma = seq(.1, 2, .1), #gamma values
              delta = c(.1, .5, 1, 2, 5, 10)) %>% #delta values
  mutate(wp = (delta*(p^gamma))/((delta*p^gamma)+(1-p)^gamma)) #images of wf

@@

* *Decision variable*: The decision variable of a prospect either counts the number of won comparisons---implying an absolute threshold---or how many more comparisons were won relative to the other prospect---implying a relative threshold.^[<!--Open footnote-->
The results are similar for absolute and relative thresholds. 
For simplification, only the results of the absolute thresholds are reported in the main text.
The results for relative thresholds are reported in the Appendix.
<!--Close footnote-->]

```{r subset-absolute, include=FALSE}
choices_absolute <- choices %>% filter(boundary == "absolute")
cpt_absolute <- cpt %>% filter(boundary == "absolute")
```

@@

@@

Overall, the results of the simulation match the predictions from Table\ \@ref(tab:tab-predictions), however, with some exceptions concerning the estimates for the parameter $\delta$ of the weighting function. 
We start by inspecting how the length of the information-processing chain changes with varying combinations of $\psi$ and $\theta$.

@@

```{r chain-length, fig.cap="(ref:chain-length)", eval = FALSE}
choices_absolute %>%
  group_by(model, s, a) %>% 
  summarize(n_med = median(n_sample)) %>% 
  ggplot(aes(x = s, y = n_med, group = as.factor(a), color = as.factor(a))) + 
  facet_wrap(~model, nrow = 1) + 
   labs(x = expression(psi),
       y = "Median of Overall Sampled Outcomes",
       color = expression(theta)) +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, .5)) +
  scale_y_continuous(limits = c(0, 120), breaks = seq(0, 150, 50)) +
  geom_point() +
  geom_line() + 
  geom_hline(yintercept = 14, linetype = "dashed", color = "gray") +
  annotate(geom="text", x=.95, y=16, label="Wulff et al. (2018)", color="gray") + 
  scale_color_viridis_d(option = "plasma") +
  theme_apa()
```

@@

```{r choice-rates, fig.cap="(ref:choice-rates)", eval = FALSE}
#prepare data

@@

(ref:weighting-parameters) Parameter Estimates of the Weighting Function

@@

label_theta <- function(string) {
  TeX(paste("$\\theta=$", string, sep = "")) #threshold parameter theta
}

# plot gamma estimates
gamma <- cpt_absolute %>%
  filter(parameter == "gamma") %>%
  ggplot(aes(s, mean, color = model)) +
  facet_wrap(~a, nrow = 2, labeller = labeller(a = as_labeller(label_theta, default = label_parsed)), scales = "free") +
  scale_x_continuous(limits = c(0,1), breaks = seq(0,1,.5)) + 
  labs(x = expression(paste("Switching Probability ", psi)), 
       y = expression(gamma),
       color = "Model") +
  geom_errorbar(aes(ymin=`2.5%`, ymax=`97.5%`)) + 
  geom_point() +
  geom_line() +
  scale_color_viridis_d(option = "viridis", begin = .3, end = .8) + 
  theme_minimal()

# plot delta estimates
delta <- cpt_absolute %>%
  filter(parameter == "delta") %>%
  ggplot(aes(s, mean, color = model)) +
  facet_wrap(~a, nrow = 2, labeller = labeller(a = as_labeller(label_theta, default = label_parsed)), scales = "free") +
  scale_x_continuous(limits = c(0,1), breaks = seq(0,1,.5)) + 
  labs(x = expression(paste("Switching Probability ", psi)), 
       y = expression(delta),
       color = "Model") +
  geom_errorbar(aes(ymin=`2.5%`, ymax=`97.5%`)) + 
  geom_point() +
  geom_line() +
  scale_color_viridis_d(option = "viridis", begin = .3, end = .8) + 
  theme_minimal()

# merge plots
ggarrange(gamma, delta, ncol = 1, nrow = 2, common.legend = TRUE, legend = "right")
```

@@

(ref:sensitivity-parameters) Parameter Estimates of the Outcome and Choice Sensitivity Parameter

```{r sensitivity-parameters, fig.cap="(ref:sensitivity-parameters)"}

# plot alpha estimates
alpha <- cpt_absolute %>%
  filter(parameter == "alpha") %>%
  ggplot(aes(s, mean, color = model)) +
  facet_wrap(~a, nrow = 2, labeller = labeller(a = as_labeller(label_theta, default = label_parsed)), scales = "free") +
  scale_x_continuous(limits = c(0,1), breaks = seq(0,1,.5)) + 
  labs(x = expression(paste("Switching Probability ", psi)), 
       y = expression(alpha),
       color = "Model") +
  geom_errorbar(aes(ymin=`2.5%`, ymax=`97.5%`)) + 
  geom_point() +
  geom_line() +
  scale_color_viridis_d(option = "viridis", begin = .3, end = .8) + 
  theme_minimal()

# plot rho estimates
rho <- cpt_absolute %>%
  filter(parameter == "rho") %>%
  ggplot(aes(s, mean, color = model)) +
  facet_wrap(~a, nrow = 2, labeller = labeller(a = as_labeller(label_theta, default = label_parsed)), scales = "free") +
  scale_x_continuous(limits = c(0,1), breaks = seq(0,1,.5)) + 
  labs(x = expression(paste("Switching Probability ", psi)), 
       y = expression(rho),
       color = "Model") +
  geom_errorbar(aes(ymin=`2.5%`, ymax=`97.5%`)) + 
  geom_point() +
  geom_line() +
  scale_color_viridis_d(option = "viridis", begin = .3, end = .8) + 
  theme_minimal()

# merge plots
ggarrange(alpha, rho, ncol = 1, nrow = 2, common.legend = TRUE, legend = "right") 
```

Figure\ \@ref(fig:function-forms) displays the resulting graphical shapes of the weighting and value function.

(ref:function-forms) Estimated Graphical Shape of the Weighting and Value Function

```{r function-forms, fig.cap="(ref:function-forms)"}

# plot weighting function

## compute decision weights

weights <- cpt_absolute %>%
  select(model, s, boundary, a, parameter, mean) %>%
  pivot_wider(names_from = parameter, values_from = mean) %>%
  select(-c(alpha, rho)) %>%
  expand_grid(p = seq(0, 1, .05)) %>%
  mutate(w = round(  (delta * p^gamma)/ ((delta * p^gamma)+(1-p)^gamma), 2))

## plot graphs

wf <- weights %>% 
  ggplot(aes(p, w, group = s, color = s)) +
  facet_wrap(~a~model, nrow = 2, labeller = labeller(a = as_labeller(label_theta, default = label_parsed))) + 
  labs(x = "Experienced Probability",
       y = expression(pi),
       color = expression(psi)) +
  scale_x_continuous(breaks = seq(0, 1, .5)) +
  scale_y_continuous(breaks = seq(0, 1, .5)) +
  geom_line(size = .5) +
  scale_color_viridis(option = "plasma") +
  theme_minimal()

# value functions

values <- cpt_absolute %>%
  select(model, s, boundary, a, parameter, mean) %>%
  pivot_wider(names_from = parameter, values_from = mean) %>%
  select(-c(gamma, delta, rho)) %>%
  expand_grid(x = seq(0, 20, .5)) %>%  
  mutate(v = round(x^alpha, 2))

vf <- values %>% 
  ggplot(aes(x, v, group = s, color = s)) +
  facet_wrap(~a~model, nrow = 2, labeller = labeller(a = as_labeller(label_theta, default = label_parsed))) + 
  labs(x = "Objective Outcome",
       y = "Subjective Value",
       color=expression(psi)) +
  scale_x_continuous(limits = c(0,20), breaks = seq(0, 20, 10)) +
  geom_line(size = .5) +
  scale_color_viridis(option = "plasma") +
  theme_minimal()

# merge plots
ggarrange(wf, vf, ncol = 1, nrow = 2, common.legend = TRUE, legend = "right")
```

# Discussion

@@
